Journey: A Study of Scalabilty of Memory Management in Linux

### 2017-05-04 paper.tex
Various changes; add spai's paragraph

In particular, we look at the case of a single process using a large amount of
memory in Linux. We run experiments to identify the scalability of the Linux
memory management system across multiple dimensions. First, we measure the
overhead of \texttt{vm\_area\_struct}, a common bookkeeping data structure in
the Linux kernel. Then, we measure the memory and processing overheads of
swapping. From these data, we extrapolate to systems running with more memory
with an eye on implications for studying large systems. While these experiments
are not comprehensive, we believe they can offer insights that are useful both
for building a research tool and for building systems.

@@

The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
intervals and sleeps in between. Also, it may be invoked if the system detects
that memory has gone below a threshold. kswapd will continue to run until enough
pages are freed to bring the free memory pool above some threashold. Then,
kswapd goes back to sleep.  One concern with systems that have a large memory
usage is that kswapd may expend significant CPU time scanning through hundreds
of thousands of pages.

@@

For each workload, we measure the latency and increase in memory usage due to
each operation in the benchmark. Notably, we do not touch the pages that we
allocate, avoiding both a page fault and the allocation of a back physical page
for the allocated virtual pages. This allows us to measure just the overhead of
memory management, as opposed to physical page allocation, swapping, page
faults, or other overheads.  Each workload contains $2^{20}$ memory management
operations.

@@

\item \textit{Random}. This workload allocates randomly the pages in a
inside of a $2^{20}$ page region.

\item \textit{Fragmented} (\texttt{frag} for short). This workload first allocates
    $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
        The resizing operation, rather than the initial allocation, is measured.
\end{itemize}

@@

A single process allocates 18GB of virtual memory space and writes one byte of
each 4KB page in the region. The same two access patterns are tested as in the
previous set of experiments (continuous and random).

@@

Figure \ref{fig:mmap_latency} shows the time taken
per \texttt{mmap} operation as it varies with the total amount memory allocated
so far by the workload.

@@

>>>>

In particular, we look at the case of a single process using a large amount of
memory in a current Linux system. We run experiments to identify the scalability
of the Linux memory management system across multiple dimensions. First, we
measure the overhead of \texttt{vm\_area\_struct}, a common bookkeeping data
structure in the Linux kernel. Then, we measure the memory and processing
overheads of swapping. From these data, we extrapolate to systems running with
more memory; many of the behaviors we identify would be aggravated by having or
using more memory.

While these experiments are not comprehensive, we believe they can offer
insights that are useful both for building a research tool and for building
systems. Also, we focus on large memory systems in general, rather than
persistent memory system in particular.

@@

The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
intervals and sleeps in between. Also, it may be invoked if the system detects
that memory has gone below a threshold. kswapd will continue to run until enough
pages are freed to bring the free memory pool above some threashold. Then,
kswapd goes back to sleep \cite{utlk}.  One concern with systems that have a large memory
usage is that kswapd may expend significant CPU time scanning through hundreds
of thousands of pages.

@@

For each workload, we measure the latency and increase in memory usage due to
each operation in the benchmark. Notably, we do not touch the pages that we
allocate, avoiding both a page fault and the allocation of a backing physical page
for the allocated virtual pages. This allows us to measure just the overhead of
memory management, as opposed to physical page allocation, swapping, page
faults, or other overheads.  Each workload contains $2^{20}$ memory management
operations.

@@

\item \textit{Random}. This workload randomly allocates the pages in a
$2^{20}$ page region.

\item \textit{Fragmented} (\texttt{frag} for short). This workload first allocates
    $2^{21}$ contiguous 4KB blocks. Then, it resizes every other block to 16KB.
        The resizing operation, rather than the initial allocation, is measured.
        By resizing every other block, rather than every block, the number of
        worst-case operations is maximized because the kernel must move each
        resized block.
\end{itemize}

@@

As in the previous set of experiments, a single process allocates 18GB of virtual memory space and writes one byte of
each 4KB page in the region. The same two access patterns are tested as in the
previous set of experiments (continuous and random).

@@

Figure \ref{fig:mmap_latency} shows the time taken
per \texttt{mmap} operation as it varies with the total amount of memory allocated
so far by the workload.

@@

(TODO: where to put this paragraph?)
A study of recent changes to memory management in Linux has found that there has
been a lot of focus on improving the scalability of data structures used in the
Linux memory manager (TODO cite: Huang et. al. ATC'16). Since systems with
persistent memory are expected to have large amounts of memory, there have been
efforts to reduce the overheads of struct page. This is because these structures
can consume several gigabytes of memory for systems with terabytes of persistent
memory (TODO cite: https://lwn.net/Articles/640113/). These structures need to
be stored in volatile memory not only because they need to be accessed quickly
for performance reasons, but also to avoid wearing out of persistent memory due
to frequent writes to this data structure. Patches proposed to the Linux kernel
have proposed getting rid of the struct page altogether, but the complexity
associated with such radical changes have hindered their acceptance by the
community. Other works such as RadixVM \cite{radixvm} have also approached
scalability with respect to the number of cores in the system.
